---
title: "EDA and Preliminary Steps"
author: "Madeleine Livaudais"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load libraries
library(ISLR2)
library(tidyverse)
library(ggplot2)
library(knitr)
library(kableExtra)
library(factoextra)    # PCA visualization - scree plots, biplots
library(cluster)       # clustering algorithms
library(tree)          # basic classification trees
library(randomForest)  # if you end up using RF
library(glmnet)        # Lasso (Ch. 6)
library(MASS)          # LDA if needed (Ch. 4)
library(boot)          # bootstrap and CV (Ch. 5)
```

## EDA 

```{r}
#| label: Read in the data and see basic structure

# Set seed for reproducability
set.seed(42)

# Inspect the data
data(Caravan)
dim(Caravan)
str(Caravan)
```

```{r}
#| label: Visualize the purchase distribution (outcome variable)

summary(Caravan$Purchase)

# Purchase distribution
ggplot(Caravan, aes(x = Purchase, fill = Purchase)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count), vjust = -0.5)) +
  labs(title = "Distribution of Caravan Insurance Purchase",
       x = "Purchase", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none")
```

Think about it from the insurance company's perspective. They're using this model to decide which customers to target with a marketing campaign. They have two ways they can make a mistake:

- Mistake 1: They predict someone will buy insurance but that person actually won't. They wasted a marketing call. Annoying but not catastrophic.

- Mistake 2: They predict someone won't buy insurance but that person actually would have. They missed a real customer. This is the costly mistake.

So what the insurance company really cares about is: of all the people who actually buy insurance, how many did our model correctly identify? That metric is called sensitivity or recall — and it's much more informative than accuracy when classes are imbalanced. In fact, accuracy can be misleading in this situation. Both of these can be calculated from a confusion matrix though.


```{r}
#| label: Predictor distributions

# Check predictor distributions
caravan_predictors <- Caravan %>% dplyr::select(-Purchase)

# Look at the spread of means and variances across variables
data.frame(
  mean = colMeans(caravan_predictors),
  variance = apply(caravan_predictors, 2, var)
) %>% summary()
```

Here we see the variance ranges from 0.0005 to 165 across predictors, making scaling essential before PCA.

```{r}
# Check sparsity - how many variables are mostly zeros?
zero_pct <- colMeans(caravan_predictors == 0) * 100

data.frame(
  variable = names(zero_pct),
  pct_zero = round(zero_pct, 1)
) %>%
  arrange(desc(pct_zero))
```

Here we show that a lot of variables are almost 99.9% zeros, meaning we get no information and they likely have 0 variance. This means that dimension reduction is very important here. 

```{r}
# Purchase rate by income level
Caravan %>%
  group_by(MINKGEM) %>%
  summarise(
    n = n(),
    purchase_rate = mean(Purchase == "Yes") * 100
  ) %>%
  ggplot(aes(x = factor(MINKGEM), y = purchase_rate)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Purchase Rate by Average Income Level",
       x = "Income Level (1=low, 6=high)",
       y = "Purchase Rate (%)") +
  theme_minimal()
```

This shows the as income increases, the purchase rate generally increases. There is a positive association between income and purchase. 

## PCA

```{r, warning=FALSE}
# Scale the predictors and run PCA
caravan_scaled <- scale(caravan_predictors)

pca_result <- prcomp(caravan_scaled, center = FALSE, scale. = FALSE)
# note: we already scaled manually, so center and scale are FALSE here

# Scree plot - how much variance does each component explain?
fviz_eig(pca_result, 
         ncp = 20,  # show first 20 components
         addlabels = TRUE,
         barfill = "steelblue",
         barcolor = "steelblue") +
  labs(title = "Scree Plot: Variance Explained by Principal Components",
       x = "Principal Component",
       y = "Percentage of Variance Explained") +
  theme_minimal()

# Cumulative variance explained
var_explained <- summary(pca_result)$importance[3,]
data.frame(
  PC = 1:40,
  cumulative_var = var_explained[1:40] * 100
) %>%
  head(40)
```

